Forecasting U.S. Food Market Prices: A Bayesian Hierarchical Approach

### 2024-12-14 paper/paper.qmd
fixing spelling errors

Food prices are a critical component of economic and social well-being, directly affecting food security, diet quality, and household expenditures. The Food-at-Home Monthly Area Prices (F-MAP) dataset, developed by the USDA Economic Research Service, offers detailed insights into food pricing trends across the United States. Covering the years 2012 to 2018, the dataset provides monthly price data for 90 food categories across 15 geographic areas, making it a valuable resource for understanding regional and temporal variations in food costs. This paper leverages the F-MAP data to analyze how regional disparities and time-based trends influence food-at-home prices and explores the role of economic factors such as purchase volume, food categories, and the Consumer Price Index (CPI) in shaping these patterns.

The primary focus of this analysis is to estimate how food categories, various economic factors, and time affect overall national food prices in the U.S. The estimand is the expected food price for a given food category and time period, conditional on factors such as purchase volume and CPI. By modeling these variations, we aim to uncover the drivers of price changes and predict trends in food costs.

@@

The Food-at-Home Monthly Area Prices (F-MAP) data product [@USDA_FMAP] is a comprehensive and detailed data product developed by the USDA Economic Research Service (ERS) that provides monthly U.S. food price data for 90 food-at-home (FAH) categories across 15 geographic areas of the United States. The dataset includes two primary price measures for each food group, geographic area, and month: (1) a mean unit value price (dollars per 100 grams) and (2) price indexes derived using advanced index formulas. These measures enable researchers to track food price trends at a granular level and compare them across geographic and temporal dimensions, while accounting for economic factors such as consumer purchasing volume, store characteristics, and inflation metrics like the Consumer Price Index (CPI). By utilizing Circana OmniMarket Core Outlets retail scanner data, the F-MAP captures detailed consumer purchasing data from over 50,000 retail stores annually, including grocery stores, supercenters, and convenience stores.

@@

The initial step involved removing unnecessary variables, retaining only those essential for analysis. This included variables such as unit_value_mean_wtd (weighted unit value), purchase_grams_wtd (weighted purchase volume), price_index_GEKS (Consumer Price Index), EFPG (food category), metroregion_name (region), and time. Rows containing missing values in any of the key variables were removed to maintain data integrity. We also filter the data to only include the national-level data, since we want to capture price trends at a national level. The dependent variable, unitValue, was log-transformed to stabilize variance and linearize relationships, to ensure no assumptions are violations in modelling. The time variable, originally in year-month format, was converted into a continuous variable representing the number of months since January 2012. This allows temporal trends to be modeled effectively as a continuous variable. The category variable was standardized and converted into factor variables, enabling the model to treat them as categorical inputs. To improve the convergence of the Bayesian model, predictors such as cpi and purchaseVolume were normalized by centering them around zero and scaling them to have a standard deviation of one.

@@

The Food-at-Home Monthly Area Prices (F-MAP) dataset is built using high-frequency retail scanner data sourced from approximately 50,000–60,000 retail establishments annually. These include grocery stores, supercenters, club stores, drug stores, and convenience stores. The scanner data capture weekly sales in nominal dollars (not adjusted for inflation) and the quantities of food items sold. Weekly sales data are aggregated into monthly intervals to align with the dataset’s temporal structure. If a sales week spans two months, the sales values and units are proportionately allocated based on the number of days in each month, ensuring temporal consistency.

@@

Through this rigorous process, real-world phenomena such as regional price variations, inflationary trends, and food category-specific dynamics are translated into structured data entries. The combination of high-frequency retail data, standardized unit values, and multilateral price indexes ensures that the F-MAP dataset is both comprehensive and precise, supporting its use in economic research and policy-making.

@@

1. **Hierarchical Structure Validity**: The assumption that categories introduce random effects is reasonable given the heterogeneity in geographic and product-specific dynamics. The assumption is based on the nested nature of the data, where observations are grouped by region and food categories. This reflects real-world dynamics, as food prices are influenced by category-specific factors (e.g., perishability, demand patterns). By introducing random effects for regions and categories, the model captures these unobserved heterogeneities effectively. For example, the average price trend in one region may systematically differ from others, and this variation is modeled through the random intercepts. Without this hierarchical framework, the model would risk oversimplifying the underlying relationships, potentially leading to biased estimates and less reliable predictions.

2. **Normality of Residuals**: The BHM assumes that the residuals (differences between observed and predicted values) are normally distributed. This assumption underpins the use of Gaussian likelihoods in Bayesian modeling and ensures that the posterior distributions are well-defined and interpretable. Normality of residuals implies that the model captures all systematic patterns in the data, leaving random noise as the primary unexplained variation. Posterior predictive checks are used to validate this assumption, comparing the observed data distribution with the model-predicted posterior distributions. If deviations from normality are detected (e.g., skewness or heavy tails), adjustments such as alternative likelihoods (e.g., Student-t distributions) can be incorporated. The normality assumption is reasonable here because the dataset represents aggregated measures (e.g., regional or category-specific prices), which tend to exhibit normal-like properties due to the Central Limit Theorem.

@@

A key strength of the model lies in its ability to capture temporal price trends while accommodating variations across food categories. For example, categories like dairy, fresh produce, and meats, known for their price volatility, are effectively modeled within this framework. The hierarchical structure allows for simultaneous analysis of overarching trends and granular category-level dynamics. This dual focus ensures that the model not only reflects overall price increases over time but also provides insights into category-specific patterns. For instance, periods of seasonal demand or supply shocks in specific categories are well-represented in the model's predictions, making it a valuable tool for understanding nuanced market dynamics.

@@

The findings from this Bayesian Hierarchical Model (BHM) offer valuable insights for policymakers and stakeholders in addressing food price stability and affordability. The model’s ability to identify temporal trends and category-specific price variations can inform targeted interventions. For instance, policymakers could design subsidies for volatile food categories, such as fresh produce or dairy, to stabilize prices and mitigate their impact on low-income households. By preemptively addressing these fluctuations, governments can help ensure food security and affordability across diverse consumer demographics.

@@

From a computational perspective, Bayesian Hierarchical Modeling (BHM) presents challenges, particularly the time required for convergence when dealing with large datasets and complex model specifications. The high-dimensional nature of hierarchical structures, can lead to long processing times and potential difficulties in achieving reliable convergence. For instance, diagnostics like high R-hat values in earlier model iterations highlight the need for adjustments, such as increasing the number of iterations, tuning parameters like adapt_delta, or simplifying priors.

@@

The findings offer actionable insights for stakeholders, particularly in predicting price volatility and understanding long-term trends. For instance, the model’s ability to identify categories with greater susceptibility to price fluctuations, such as fresh produce and dairy, provides a foundation for targeted interventions. Furthermore, the strong predictive accuracy demonstrated by the model underscores its potential as a tool for preempting affordability crises, stabilizing food prices, and informing policy decisions.

@@

Bilateral indexes with a fixed base period can become less representative of the cost of food as the indexes move further away from the base, due to the effects of product turnover, as products are discontinued and new products are introduced. Although bilateral price indexes can be updated using chained indexes, which capture product substitution, chained indexes are subject to drifting. Chain drift is a phenomenon in which the price index drifts lower even as item-level prices return to their base levels. Multilateral indexes are fully transitive and free of chain drift. While chain drift is possible in rolling-window multilateral indexes, if a wide window length is chosen, the rolling-window index will be largely free of drift despite not being fully transitive. The 1-year rolling window used in the F-MAP GEKS has been found to be sufficient to remove chain drift caused by high-frequency data and seasonal variation in variety.

>>>>

Food prices are a critical component of economic and social well-being, directly affecting food security, diet quality, and household expenditures. The Food-at-Home Monthly Area Prices (F-MAP) dataset, developed by the USDA Economic Research Service, offers detailed insights into food pricing trends across the United States. Covering the years 2012 to 2018, the dataset provides monthly price data for 90 food categories across 15 geographic areas, making it a valuable resource for understanding regional and temporal variations in food costs. This paper leverages the F-MAP data to analyze how regional disparities, and time-based trends influence food-at-home prices and explores the role of economic factors such as purchase volume, food categories, and the Consumer Price Index (CPI) in shaping these patterns.

The primary focus of this analysis is to estimate how food categories, various economic factors, and time affect overall national food prices in the U.S. The estimand is the expected food price for a given food category and time period, conditional on factors such as purchase volume and CPI. By modelling these variations, we aim to uncover the drivers of price changes and predict trends in food costs.

@@

The Food-at-Home Monthly Area Prices (F-MAP) data product [@USDA_FMAP] is a comprehensive and detailed data product developed by the USDA Economic Research Service (ERS) that provides monthly U.S. food price data for 90 food-at-home (FAH) categories across 15 geographic areas of the United States. The dataset includes two primary price measures for each food group, geographic area, and month: (1) a mean unit value price (dollars per 100 grams) and (2) price indexes derived using advanced index formulas. These measures enable researchers to track food price trends at a granular level and compare them across geographic and temporal dimensions, while accounting for economic factors such as consumer purchasing volume, store characteristics, and inflation metrics like the Consumer Price Index (CPI). By utilizing Circana OmniMarket Core Outlets retail scanner data, the F-MAP captures detailed consumer purchasing data from over 50,000 retail stores annually, including grocery stores, supercentres, and convenience stores.

@@

The initial step involved removing unnecessary variables, retaining only those essential for analysis. This included variables such as unit_value_mean_wtd (weighted unit value), purchase_grams_wtd (weighted purchase volume), price_index_GEKS (Consumer Price Index), EFPG (food category), metroregion_name (region), and time. Rows containing missing values in any of the key variables were removed to maintain data integrity. We also filter the data to only include the national-level data, since we want to capture price trends at a national level. The dependent variable, unitValue, was log-transformed to stabilize variance and linearize relationships, to ensure no assumptions are violations in modelling. The time variable, originally in year-month format, was converted into a continuous variable representing the number of months since January 2012. This allows temporal trends to be modelled effectively as a continuous variable. The category variable was standardized and converted into factor variables, enabling the model to treat them as categorical inputs. To improve the convergence of the Bayesian model, predictors such as cpi and purchaseVolume were normalized by centering them around zero and scaling them to have a standard deviation of one.

@@

The Food-at-Home Monthly Area Prices (F-MAP) dataset is built using high-frequency retail scanner data sourced from approximately 50,000–60,000 retail establishments annually. These include grocery stores, supercentres, club stores, drug stores, and convenience stores. The scanner data capture weekly sales in nominal dollars (not adjusted for inflation) and the quantities of food items sold. Weekly sales data are aggregated into monthly intervals to align with the dataset’s temporal structure. If a sales week spans two months, the sales values and units are proportionately allocated based on the number of days in each month, ensuring temporal consistency.

@@

Through this rigorous process, real-world phenomena such as regional price variations, inflationary trends, and food category-specific dynamics are translated into structured data entries. The combination of high-frequency retail data, standardized unit values, and multilateral price indexes ensures that the F-MAP dataset is both comprehensive and precise, supporting its use in economic research and policymaking.

@@

1. **Hierarchical Structure Validity**: The assumption that categories introduce random effects is reasonable given the heterogeneity in geographic and product-specific dynamics. The assumption is based on the nested nature of the data, where observations are grouped by region and food categories. This reflects real-world dynamics, as food prices are influenced by category-specific factors (e.g., perishability, demand patterns). By introducing random effects for regions and categories, the model captures these unobserved heterogeneities effectively. For example, the average price trend in one region may systematically differ from others, and this variation is modelled through the random intercepts. Without this hierarchical framework, the model would risk oversimplifying the underlying relationships, potentially leading to biased estimates and less reliable predictions.

2. **Normality of Residuals**: The BHM assumes that the residuals (differences between observed and predicted values) are normally distributed. This assumption underpins the use of Gaussian likelihoods in Bayesian modelling and ensures that the posterior distributions are well-defined and interpretable. Normality of residuals implies that the model captures all systematic patterns in the data, leaving random noise as the primary unexplained variation. Posterior predictive checks are used to validate this assumption, comparing the observed data distribution with the model-predicted posterior distributions. If deviations from normality are detected (e.g., skewness or heavy tails), adjustments such as alternative likelihoods (e.g., Student-t distributions) can be incorporated. The normality assumption is reasonable here because the dataset represents aggregated measures (e.g., regional or category-specific prices), which tend to exhibit normal-like properties due to the Central Limit Theorem.

@@

A key strength of the model lies in its ability to capture temporal price trends while accommodating variations across food categories. For example, categories like dairy, fresh produce, and meats, known for their price volatility, are effectively modelled within this framework. The hierarchical structure allows for simultaneous analysis of overarching trends and granular category-level dynamics. This dual focus ensures that the model not only reflects overall price increases over time but also provides insights into category-specific patterns. For instance, periods of seasonal demand or supply shocks in specific categories are well-represented in the model's predictions, making it a valuable tool for understanding nuanced market dynamics.

@@

The findings from this Bayesian Hierarchical Model (BHM) offer valuable insights for policymakers and stakeholders in addressing food price stability and affordability. The model’s ability to identify temporal trends and category-specific price variations can inform targeted interventions. For instance, policymakers could design subsidies for volatile food categories, such as fresh produce or dairy, to stabilize prices and mitigate their impact on low-income households. By pre-emptively addressing these fluctuations, governments can help ensure food security and affordability across diverse consumer demographics.

@@

From a computational perspective, Bayesian Hierarchical Modelling (BHM) presents challenges, particularly the time required for convergence when dealing with large datasets and complex model specifications. The high-dimensional nature of hierarchical structures, can lead to long processing times and potential difficulties in achieving reliable convergence. For instance, diagnostics like high R-hat values in earlier model iterations highlight the need for adjustments, such as increasing the number of iterations, tuning parameters like adapt_delta, or simplifying priors.

@@

The findings offer actionable insights for stakeholders, particularly in predicting price volatility and understanding long-term trends. For instance, the model’s ability to identify categories with greater susceptibility to price fluctuations, such as fresh produce and dairy, provides a foundation for targeted interventions. Furthermore, the strong predictive accuracy demonstrated by the model underscores its potential as a tool for pre-empting affordability crises, stabilizing food prices, and informing policy decisions.

@@

Bilateral indexes with a fixed base period can become less representative of the cost of food as the indexes move further away from the base, due to the effects of product turnover, as products are discontinued, and new products are introduced. Although bilateral price indexes can be updated using chained indexes, which capture product substitution, chained indexes are subject to drifting. Chain drift is a phenomenon in which the price index drifts lower even as item-level prices return to their base levels. Multilateral indexes are fully transitive and free of chain drift. While chain drift is possible in rolling-window multilateral indexes, if a wide window length is chosen, the rolling-window index will be largely free of drift despite not being fully transitive. The 1-year rolling window used in the F-MAP GEKS has been found to be sufficient to remove chain drift caused by high-frequency data and seasonal variation in variety.
